Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict

### 2020-09-21 paper.Rmd
final monday change
andy check several places herein where I ask you to clarify. Over to you.

---
title: "Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com.]
 - Richard K. Morgan^[Varieties of Democracy Institute, University of Gothenburg, rick.morgan2@gmail.com.]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com. Corresponding author.]
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: John Ahlquist, Cassy L. Dorff, and Shahryar Minhas both provided helpful feedback on this project. We are especially grateful to Paul Huth for his comments and guidance. All the code and several additional analyses can be found at our replication archive at https://github.com/andybega/Blair-Sambanis-replication.
abstract: We examine the research protocols in Blair and Sambanis (2020).  We find that there are several important mistakes and research decisions that determine their results. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.  Their model is s outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.  
output: 
  bookdown::pdf_document2:
    keep_tex: true
    keep_md: true
    toc: false
#header-includes:
#  - \usepackage{setspace}\doublespacing
bibliography: "./references.bib"
nocite: |
  @cederman:weidmann:2017
---

@@

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately". Indeed, B&S argue that this theoretical model outperforms four other "more mechanical" alternatives.They also claim that the integration of structure with process is better than alternaties over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.\footnote{\cite{PITF},\cite{ICEWS},\cite{VIEWS}\cite{others}} B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

We believe that B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set. This appears to be an easily corrected coding error on the part of B&S.

@@

This description suggests that B&S intended to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they both utilize the *exact same training data*, which consist of the full training data from all other models. In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split. 
The model specifications are also identical; i.e., they use the same *x* variables and the same random forest hyper-parameter settings. The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself. If we ran both with the same random seed, they would be identical in every respect, producing to forests of identical decision trees and, thus, identical predictions.

@@

B&S present a confusion matrix to score their 2016-H1 forecasts in Table 4. Although the forecasts are for the probability of civil war _onset_, in the replication code, they are actually scored using _incidence_ of civil war. By definition the incidence of civil war is much more common than onset, which is only coded for the starting year of a conflict. When forecasting rare events, it is common to have many false positives, i.e. instances where we predict onset but not onsets occur. _Ex ante_, using incidence rather than onset to score onset forecasts should decrease the number of false positives and correspondingly increase the number of true positives (good predictions) and false negatives (missed onsets).

@@

What impact did the ROC smoothing have overall on the evaluated performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B\&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B\&S Table 2)?  Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values. Each point corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed} (the colors match those in Figure \ref{fig:rocs}). The vertical bar in each plot marks the average effect of smoothing on AUC-ROC. For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible in average. The Escalation model _always_ benefits from smoothing, with an average improvement on the order of 0.05. This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the incorrect result B&S report, namely that the escalation model is generally superior.

@@

B&S interpret their results as follows, on page 20 and we comment on the conclusions seratim:^[We list the "Overall, ..." interpretation out-of-order for clarity.]

@@

The correct confusion matrices when using observed onset (or the lack of it) are shown in the second part of Table \ref{tab:table4}. In the default "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1. Thus, the recall value is undefined, while the precision is 0/30 = 0, compared to reported recall and precision values of 13/15 = 0.87 and 13/30 = 0.43. The alternative coding ("Assuming Change") produces two civil war onsets. Recall is 1.0 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.^[Another, minor issue is related to using a lead version of the DV. With the lead version of the DV, "incidence_civil_ns_plus1", which is what the models are predicting, the predicted value for 2016-H1 indicates the risk of civil war onset in 2016-H2. In the Table 4 script referenced above, the 2016-H1 predictions (for 2016-H2) are assessed using the raw DV, "incidence_civil_ns", not the lead version. Essentially, the forecasts for 2016-H2 are assessed using observed outcomes for 2016-H1. In this case, it doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.]

@@

B\&S advocate the use of theory to guide prediction. But "theory" is not defined by B\&S, and may an ambiguous and undefined concept. `Theory'' has many interpretations in social science.  For some---the EITM movement, for example [@aldrich:etal:2008]---theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.  The additional statements can be subjected to empirical scrutiny.  An example might be William Riker's theory of coalitions [@riker:1962] which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically. A more recent example of strong theory is found in *The Political Logic of Survival* [@bdm:etal:2005]. For the purposes of discussion we call this strong theory.

@@

At this level of detail the theoretical model is more about choosing the correct variables and this is the contrast between theoretical prediction models and atheoretical machine learning that B&S focus on. But this dichotomy is misguided. The data that are used in a predictive model have to come from somewhere. Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful. It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well. Rather, we posit that most of the cases B&S might characterize as a-theoretical enter the "theoretical" , with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}

Another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility in regard to functional form. Theory should not only tell us what factors matter, but also how they relate to each other. We think that few would be strongly committed to the idea that the linear additive equations, maybe with the occasional polynomial or multiplicative interaction terms, that dominate published regression analyses, are *the* correct functional specification. Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low? Are whatever those relationships are linear in the log-odds? Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships. B&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives.

@@

Blair & Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain.  As social science becomes more adept at predictive analysis, this will doubtless be of increasing importance.  However, these comparisons must be made carefully to ensure those correct inferences are drawn.  We continue to think that weak theory is overrated [@ward:2016], and that machine learning and big data will allow us to learn new things. It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation.  Social science has leaned on theory for (at least) seventy years, but it might be time to find useful ways to incorporate some of the recently developed methods in the arena of machine learning as well.

>>>>

---
title: "Reassessing the Role of Theory and Machine Learning in Forecasting Civil Conflict"
author: 
 - Andreas Beger^[Predictive Heuristics, adbeger@gmail.com.]
 - Richard K. Morgan^[Varieties of Democracy Institute, University of Gothenburg, rick.morgan2@gmail.com.]
 - Michael D. Ward^[Predictive Heuristics, Duke University, University of Washington, michael.don.ward@gmail.com. Corresponding author.]
date: "`r format(Sys.time(), '%d %B %Y')`"
thanks: John Ahlquist, Cassy L. Dorff, and Shahryar Minhas both provided helpful feedback on this project. We are especially grateful to Paul Huth for his comments and guidance. All the code and several additional analyses can be found at our replication archive at https://github.com/andybega/Blair-Sambanis-replication.
abstract: We examine the research protocols in Blair and Sambanis (nd).  We find that there are several important mistakes and research decisions that determine their results. Fixing these mistakes results in a reversal of their claim that theory based models of escalation are better at predicting onsets of civil war than other kinds of models.  Their model is s outperformed by several of the ad hoc, putatively non-theoretical models they devise and examine.  
output: 
  bookdown::pdf_document2:
    keep_tex: true
    keep_md: true
    toc: false
#header-includes:
#  - \usepackage{setspace}\doublespacing
bibliography: "./references.bib"
nocite: |
  @cederman:weidmann:2017
---

@@

B\&S claim (page 3) to show that a model informed by procedural theories of escalation and de-escalation can predict the onset of civil wars "remarkably accurately". Indeed, B&S argue that this theoretical model outperforms four other "more mechanical" alternatives.They also claim that the integration of structure with process is better than alternatives over short forecasting windows. Third, they preregistered the list of thirty countries that have the highest risk of civil war onset. They claim that such prospective predictions are rare in the literature when, in fact, they have been routine for many years with several prominent projects.\footnote{\cite{PITF},\cite{ICEWS},\cite{VIEWS}\cite{others}} B\&S claim to be unique in assessing these forecasts. A qualitative analysis of their predictions allows them to conclude that their model is robust. We will return to their analysis later, after correcting the procedural mistakes we found in their research process.

@@

We believe that B&S intend that the escalation model's predictions for the test set are weighted by the PITF model predictions for the test set. However, they actually weight the \textbf{test} set predictions using the PITF model predictions for the \textbf{training} set. This appears to be an easily corrected coding error on the part of B&S.  ANDY ARE YOU SURE THIS PARAGRAPH IS CORRECT?  DON'T YOU WANT TO WEIGHT THE TEST BY THE TRAINING?

@@

This description suggests that B&S intended to run two separate random forest models, one each on the low- and high-risk training data splits. The replication code does indeed run two separate random forecasts, but they both utilize the *exact same training data*, which consist of the full training data from all other models. In short, rather than using only data for high-risk countries to train their split-sample model for high-risk cases, the data they use captures both high- and low-risk countries, similarly for the training set for their low-risk split. 
The model specifications are also identical; i.e., they use the same *x* variables and the same random forest hyper-parameter settings. The *only* difference in the models as they are implemented in the B&S replication code is due to the non-deterministic nature of the random forest model itself. If we ran both with the same random seed, they would be identical in every respect, producing two forests of identical decision trees and, thus, identical predictions.

@@

B&S present a confusion matrix to score their 2016-H1 forecasts in Table 4. Although the forecasts are for the probability of civil war _onset_, in the replication code, they are actually scored using _incidence_ of civil war. By definition the incidence of civil war is much more common than onset, which is only coded for the starting year of a conflict. When forecasting rare events, it is common to have many false positives, i.e. instances where we predict onset but no onsets occur. _Ex ante_, using incidence rather than onset to score onset forecasts should decrease the number of false positives and correspondingly increase the number of true positives (good predictions) and false negatives (missed onsets).

@@

What impact did the ROC smoothing have overall on the evaluated performance of the Escalation model relative to other ICEWS models (our Table \ref{tab:table1-nosmooth}, B\&S Table 1) and structural extensions (our Table \ref{tab:table2-fixed}, B\&S Table 2)?  Figure \ref{fig:benefit-plot} shows the changes in AUC-ROC values had we used smoothed ROC curves to calculate the AUC-ROC values. Each point corresponds to the change in AUC-ROC values for one of the models in the cells in Tables \ref{tab:table1-nosmooth} and \ref{tab:table2-fixed} (the colors match those in Figure \ref{fig:rocs}). The vertical bar in each plot marks the average effect of smoothing on AUC-ROC. For all alternative models, from Quad to "PITF Only", smoothing sometimes hurts and sometimes benefits, but the overall impact is negligible on average. The Escalation model _always_ benefits from smoothing, with an average improvement on the order of 0.05. This is sufficient to push the Escalation model ahead of the alternative models, and accounts for the incorrect result B&S report, namely that the escalation model is generally superior.

@@

B&S interpret their results as follows (page 20) and we comment on the conclusions seratim:^[We list the "Overall, ..." interpretation out-of-order for clarity.]

@@

ANDY, THIS SECTION 4.6 AND THE TABLE 4 ARE VERY CONFUSING. CAN YOU TRY TO CLEAN IT UP?
The correct confusion matrices when using observed onset (or the lack of it) are shown in the second part of Table \ref{tab:table4}. In the default "Assuming Persistence" coding, there are no civil war onsets in the data for 2016-H1. Thus, the recall value is undefined, while the precision is 0/30 = 0, compared to reported recall and precision values of 13/15 = 0.87 and 13/30 = 0.43. The alternative coding ("Assuming Change") produces two civil war onsets. Recall is 1.0 compared to 14/16 = 0.88 before, and precision is 2/30 = 0.07 instead of 14/30 = 0.47.^[Another, minor issue is related to using a lead version of the DV. With the lead version of the DV, "incidence_civil_ns_plus1", which is what the models are predicting, the predicted value for 2016-H1 indicates the risk of civil war onset in 2016-H2. In the Table 4 script referenced above, the 2016-H1 predictions (for 2016-H2) are assessed using the raw DV, "incidence_civil_ns", not the lead version. Essentially, the forecasts for 2016-H2 are assessed using observed outcomes for 2016-H1. In this case, it doesn't make a difference since both the raw DV and lead version for 2016-H1 do not have any positive values.]

@@

B\&S advocate the use of theory to guide prediction. But "theory" is not defined by B\&S and may be an ambiguous and undefined concept. `Theory'' has many interpretations in social science.  For some---the EITM movement, for example [@aldrich:etal:2008]---theory is a system of internally consistent, logically connected propositions from which additional statements may be deduced.  The additional statements can be subjected to empirical scrutiny.  An example might be William Riker's theory of coalitions [@riker:1962] which posits propositions in game theoretic form from which the size principal of minimal winning coalitions may be deduced and further examined empirically. A more recent example of strong theory is found in *The Political Logic of Survival* [@bdm:etal:2005]. For the purposes of discussion we call this strong theory.

@@

At this level of detail the theoretical model is more about choosing the correct variables and this is the contrast between theoretical prediction models and atheoretical machine learning that B&S focus on. But this dichotomy is misguided. The data that are used in a predictive model have to come from somewhere. Even if the inclusion of some factors is not explicitly justified, it is hardly the case that one assembles data without the expectation that it will be useful. It is not a coincidence that nobody uses 1,180 variable strawman models whose variable vectors predominantly consist of 0's, even if such a model coincidentally ends up performing well. Rather, we posit that most of the cases B&S might characterize as atheoretical enter the "theoretical" , with some awareness of existing arguments and work that is used to select data sources for inclusion in a dataset.\footnote{While we do not undertake this, there are a variety of procedures that one might use for variable choices, ranging from staring at lists of variables to machine learning approaches.}

Another way in which we could contrast theoretical from atheoretical prediction models is in terms of their flexibility in regard to functional form. Theory should not only tell us what factors matter, but also how they relate to each other. We think that few would be strongly committed to the idea that the linear additive equations, maybe with the occasional polynomial or multiplicative interaction terms, that dominate published regression analyses, are *the* correct functional specification. Is the danger posed by low-level violence the same when tensions are high, with high demands and high non-violent repression, as when tensions are low? Are those relationships linear in the log-odds? Some machine learning models, random forests included, are flexible in this regard and to some extent capable of learning non-linear and interactive relationships. B&S implicitly concede on this point since they use random forests for both their Escalation model and the atheoretical alternatives.

@@

Blair & Sambanis have focused attention on comparing forecasting models and forecasts in the civil war domain.  As social science becomes more adept at predictive analysis, this will doubtless be of increasing importance.  However, these comparisons must be made carefully to ensure those correct inferences are drawn.  We continue to think that weak theory is overrated [@ward:2016], and that machine learning and big data will allow us to learn new things. It will be interesting to see how the evidence for these claims is adjudicated with additional usage and careful evaluation.  Social science has leaned on theory for (at least) seventy years, but it might be time to find useful ways to incorporate some of the recently developed methods in the arena of machine learning and modern statistics as well.
